---
title: "Customer Satisfaction MOPC"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "centre")
```

```{r}
library(tidyverse)
```

```{r}
data<-read.csv("C://Users//adm//Desktop//HCQ//Research dept//MedicalOutpatientCli_DATA_LABELS_2022-03-09_1207.csv")
```

# Exploratory Data Analysis.

## How  you would rate the following aspects in the clinic today.

###  Cleaniliness.
```{r fig.cap= "figure 1.1"}
ggplot(data = data,aes(x=factor(Cleanliness),
                       y=prop.table(stat(count)),
                         fill=factor(Cleanliness),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_viridis_d(begin=0, end = .95,option='D')+
  labs(x="Cleanliness rating",y="Percentage proportion",title = "Categorizationof cleanliness in the hospital")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))
```

The above plot is very effective in showing how cleanliness was rated by different customers in the hospital.A majority of the customers that is, 37.61% rated it as **very good**, following closely was **excellent** at 31.19 % . **Neutral** came in third at 0.61% and finally those who rated it as fair were 0.31%.There was no customer who rated the hospital cleanliness as poor.

### Customer friendliness.
```{r fig.cap="Figure 1.2"}
ggplot(data = data,aes(x=factor(Customer.friendliness),
                       y=prop.table(stat(count)),
                         fill=factor(Customer.friendliness),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Dark2")+
  theme_light()+
  labs(x="Customer friendliness rating",y="Percentage proportion",title = " Customer friendliness in the hospital")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))
```
```{r eval=FALSE}
table(data$Customer.friendliness)
```

Majority of the respondents 34.25% (n=112) termed customer friendliness in the hospital as *very good*,*good* was at 33.94% (n=111). Those who felt customer friendliness was *excellent* accounted for 22.94% (n=75). *Neutral*,*fair* and *poor* came in forth ,fifth and sixth at 4.53%(n=15),2.45% (n=8) and 1.83% (n=6) respectively.

### Security
```{r fig.cap= "Figure 1.3"}
ggplot(data = data,aes(x=factor(Security),
                       y=prop.table(stat(count)),
                         fill=factor(Security),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Accent")+
  theme_light()+
  labs(x="Security rating",y="Percentage proportion",title = " Security in the hospital")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))
```
```{r eval=FALSE}
table(data$Security)
```
In regards to the question of security within the hospital a large number of our respondents 35.78% (n=117) felt that security was **good**. 31.8% (n=104) rated it as **very good**,those that felt it was **excellent** accounted for 22.94% (n=75).**Neutral**, **fair** and **poor** followed each other closely at 2.75%,2.45% and 2.14% respectively. 2.14% (n=7) did not respond to this question.

### Privacy.
```{r fig.cap= "Figure 1.4"}
ggplot(data = data,aes(x=factor(Privacy),
                       y=prop.table(stat(count)),
                         fill=factor(Privacy),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Paired")+
  theme_update()+
  labs(x="Privacy rating",y="Percentage proportion",title = " Privacy rating in the hospital")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))
```
```{r eval=FALSE}
table(data$Privacy)
```


A large percentage of  respondents 43.4% (n=142) rated customer privacy as **excellent**,24.8% felt it was **very good** (n=81). Those who termed customer privacy as *good* accounted for 21.7% (n=71).**Neutral**and **fair**  tied at 4% and *poor* came in last at 2.1%(n=7).


## Did the doctor talk to you in a courteous manner? 

```{r fig.cap="figure 2"}
ggplot(data = data,aes(x=factor(Did.the.doctor.talk.to.you.in.a.courteous.manner.),
                       y=prop.table(stat(count)),
                         fill=factor(Did.the.doctor.talk.to.you.in.a.courteous.manner.),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Accent")+
  theme_light()+
  labs(x="Did the doctor talk to you in a courteous manner?",y="Percentage proportion",title = " ")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))
```
```{r eval=FALSE}
table(data$Did.the.doctor.talk.to.you.in.a.courteous.manner.)

```

From figure 2 above it is evident that 99% (n=324)of the respondents felt that the the doctor talked to them in a courteous manner. Only 1% (n=3) of the respondents felt otherwise.

## During triage were you handled courteously? 
```{r fig.cap= "Figure 3"}
ggplot(data = data,aes(x = factor(During.triage.were.you.handled.courteously.),
                       y = prop.table(stat(count)),
                       fill= factor(During.triage.were.you.handled.courteously.),
                       label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  scale_y_continuous(labels = scales::percent)+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  labs(x = "During triage were you handled courteously",y = "percentage proportion",title = "Whether customer was handled courteously ")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"),)+
  scale_fill_brewer(palette = "Dark2")+
  scale_x_discrete(labels = c("Did not respond","No","Yes"))

```
```{r eval=FALSE}
table(data$During.triage.were.you.handled.courteously.)

```

Majority of the respondents, 92.7% (n= 303) in the study question (*During triage were you handled courteously?*) were of the opinion that they were handled courteously during the triage.2.1% (n=7) felt that they were not handled courteously and 5.2% (n=17) did not  give their opinion in regards to this question.



## Did the nurse explain the procedure above before performing it? 
```{r fig.cap="Figure 4" }
ggplot(data = data,aes(x=factor(Did.the.nurse.explain.the.procedure.above.before.performing.it.),
                       y=prop.table(stat(count)),
                         fill=factor(Did.the.nurse.explain.the.procedure.above.before.performing.it.),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Set2")+
  theme_light()+
  labs(x="Did the nurse explain the procedure above before performing it?",y="Percentage proportion",title = "Whether the Nurse explained procedure \n before performing it ")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_x_discrete(labels = c("Did not respond","No","Yes"))
```
```{r eval=FALSE}
table(data$Did.the.nurse.explain.the.procedure.above.before.performing.it.)
```
Figure 4 above shows that 64% (n=208) of the respondents were explained to the above procedure by the nurse before it was performed.31% (n=102) were not explained to the procedure and 5% (n=17) did not respond to this question.

## Were the Health information officers courteous?
```{r fig.cap= "figure 5" }
ggplot(data = data,aes(x=factor(Were.the.Health.information.officers.courteous.),
                       y=prop.table(stat(count)),
                         fill=factor(Were.the.Health.information.officers.courteous.),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Set3")+
  theme_light()+
  labs(x="Were the Health information officers courteous?",y="Percentage proportion",title = "Whether the Health information officers courteous")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_x_discrete(labels = c("Did not respond","No","Yes"))
```
```{r eval=FALSE}
table(data$Were.the.Health.information.officers.courteous.)
```


92.4% (n=302) of the respondents felt that the Health information officers courteous, 7% (n=23) were of a contrary opinion and 0.6% (n=2) did not respond to this question.

## **How do you rate the length of the following services?**
### Payment

```{r fig.cap="Figure 6.1" }
ggplot(data = data,aes(x=factor(Payment),
                       y=prop.table(stat(count)),
                         fill=factor(Payment),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Dark2")+
  theme_light()+
  labs(x="Customer rating of the length of paymet",y="Percentage proportion",title = "Rating of the length of paymet")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_x_discrete(labels = c("Did not respond","Expected","Too Long"))
```
```{r eval=FALSE}
table(data$Payment)
```

Majority  of the respondents 87% (n=283) were of the opinion that time length at payment met their expectation and 13% (n=43) felt it was too long .This question a 100% response

### Health information
```{r fig.cap= "Figure 6.2"}
ggplot(data = data,aes(x=factor(Health.information),
                       y=prop.table(stat(count)),
                         fill=factor(Health.information),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Paired")+
  theme_light()+
  labs(x="Customer rating of Health information",y="Percentage proportion",title = "Rating of Health information")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_x_discrete(labels = c("Did not respond","Expected","Too Long"))
```
```{r eval=FALSE}
table(data$Health.information)
```
70% (n=230) of the respondents were of the opinion that time spent at health information was as they expected and 29% (n=95) felt it was too long. Only 1% of the respondents did not respond to this question.


### Triage
```{r fig.cap="Figure 6.3"}
ggplot(data = data,aes(x=factor(Triage),
                       y=prop.table(stat(count)),
                         fill=factor(Triage),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Dark2")+
  theme_light()+
  labs(x="Customer rating of Triage",y="Percentage proportion",title = "Rating of Triage")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_x_discrete(labels = c("Did not respond","Expected","Too Long"))
```
```{r eval=FALSE}
table(data$Triage)
```

Majority of the respondents, 92.7% (n= 303) in the study question (*rate of length at triage*) were of the opinion that the time spent at triage was as expected.1.8% (n=6) felt that it was too long  and 5.5% (n=18) did not  give their opinion in regards to this question.

### Doctor's review

```{r fig.cap="Figure 6.4"}
ggplot(data = data,aes(x=factor(Doctor.s.review),
                       y=prop.table(stat(count)),
                         fill=factor(Doctor.s.review),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_brewer(palette = "Accent")+
  theme_light()+
  labs(x="Customer rating of Doctor's review",y="Percentage proportion",title = "Rating of Doctor's review")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_x_discrete(labels = c("Did not respond","Expected","Too Long"))
```
```{r eval=FALSE}
table(data$Doctor.s.review)
```

Of all the respondents 85% (n= 277)  were of the opinion that the time spent at at doctors review was as expected.15% (n=49) were of the opinion it was too long  and 0% (n=1) did not  give their opinion in regards to this question.

```{r fig.cap="Figure 6.5"}
ggplot(data = data,aes(x=factor(Appointment.booking),
                       y=prop.table(stat(count)),
                         fill=factor(Appointment.booking),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_viridis_d()+
  theme_light()+
  labs(x="Customer rating of Appointment booking",y="Percentage proportion",title = "Rating of Appointment booking")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_x_discrete(labels = c("Did not respond","Expected","Too Long"))
```

```{r eval=FALSE}
table(data$Appointment.booking)
```
Majority of the respondents, 61% (n= 201) in the study question (*rate of length at triage*) were of the opinion that the time spent at triage was as expected.1.8% (n=6) felt that it was too long  and 5.5% (n=18) did not  give their opinion in regards to this question.



## In future, would you seek health care services in this hospital?

```{r fig.cap="Figure 7"}
ggplot(data = data,aes(x=factor(In.future..would.you.seek.health.care.services.in.this.hospital.),
                       y=prop.table(stat(count)),
                         fill=factor(In.future..would.you.seek.health.care.services.in.this.hospital.),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_viridis_d(option = "D",end = .9)+
  theme_light()+
  labs(x="In future, would you seek health care services in this hospital",y="Percentage proportion",title = "In future, would you seek \n health care services in this hospital")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))
  #scale_x_discrete(labels = c("Did not respond","Expected","Too Long"))
```
94% of the respondents felt that in future, would you seek health care services in this hospital while 6% felt they would not seek medication at KNH again




## Would you refer a friend/relative to Kenyatta National Hospital?

```{r fig.cap="Figure 8"}
ggplot(data = data,aes(x=factor(Would.you.refer.a.friend.relative.to.Kenyatta.National.Hospital.),
                       y=prop.table(stat(count)),
                         fill=factor(Would.you.refer.a.friend.relative.to.Kenyatta.National.Hospital.),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_viridis_d(option = "B",begin = 0.3,end = .67)+
  theme_light()+
  labs(x="Would you refer a friend/relative to Kenyatta National Hospital",title = "Would you refer a friend/relative \n to Kenyatta National Hospital")+
  theme(legend.position = "none",plot.title = element_text(face = "bold"))
```
```{r eval=FALSE}
table(data$Would.you.refer.a.friend.relative.to.Kenyatta.National.Hospital.)
```
97% (n=316) of the respondents felt they would you refer a friend/relative to Kenyatta National Hospital and 3% (n=11) felt they would not.