---
title: 'Reduction of waiting time for patients awaiting echo '
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
library(tidyverse)
```

```{r}
data_11<- data.frame(quiz1a = c(rep(1,14),rep(2,14),"NA"),
                   quiz1b = c(rep(1,27),rep(2,2)),
                   quiz1c = c(rep(1,25),rep(2,4)),
                   quiz1d = c(rep(1,24),rep(2,5)),
                   quiz1e = c(rep(1,21),rep(2,8)),
                   quiz1f = c(rep(1,28),rep(2,1)),
                   quiz2a = c(rep(1,27),rep(2,2)),
                   quiz2b = c(rep(1,25),rep(2,4)),
                   quiz2c = c(rep(1,23),rep(2,4),"NA","NA"),
                   quiz2d = c(rep(1,20),rep(2,9)),
                   quiz2e = c(rep(1,19),rep(2,9),"NA"),
                   quiz2f = c(rep(1,23),rep(2,5),"NA"),
                   quiz3a = c(rep(1,29),rep(2,0)),
                   quiz3b = c(rep(1,28),rep(2,1)),
                   quiz3c = c(rep(1,25),rep(2,3),"NA"))
```

```{r}
data_11 %>%
  gather(key="Questions",value, quiz1a:quiz1f) -> p
pp<-subset(p, select = c('Questions' , 'value'))
pp$Questions<-"policy"
pp<-pp[pp$value==1,]

data_11 %>%
  gather(key="Questions",value, quiz2a:quiz2f) -> h
hh<-subset(h, select = c('Questions' , 'value'))
hh$Questions<-"people"
hh<-hh[hh$value==1,]

data_11 %>%
  gather(key="Questions",value, quiz3a:quiz3c) -> i
ii<-subset(i, select = c('Questions' , 'value'))
ii$Questions<-"machine"
ii<-ii[ii$value==1,]

n<-rbind(pp,hh,ii)


ggplot(data=n,aes(x=factor(Questions),y=prop.table(stat(count)),
                  fill = factor(Questions),
                  label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  scale_y_continuous(labels = scales::percent)+
  geom_text(stat = 'count',
            vjust= -0.5,
            size= 3)+
   scale_fill_brewer(palette = "Dark2")+
  #scale_x_discrete( labels =c("Yes","No"))+
  labs(x = "Causes of delay",y = "Percentage Proportion",title = "Reduction of waiting time \n for patients awaiting echo")+
  theme(legend.position = "none",plot.title = element_text(face="bold"))
```

<br> 

Analysis of data collected on some of the issues the pediatrics Department had identified as causes of delay in echo process had the following results.

**Policy** had the greatest impact on the question under study at **38.83%**.As seen from the above graph **people** came in as the second cause of the delays at an average percentage of **38.27%**.The least  cause of delay was machine at an average percentage of **22.19%**.

The questions that were open for study  are listed below:

**POLICY**

a) I am not aware of the echo booking and procedure work instruction/SOP

b) There is delay in referring patients from peripheral health care facilities hence they come as emergencies and affect those already waiting

c) Multiple investigations required before echo( CXR,ECG etc) leading to delay.

d) Multiple consultations.

e) There is no standardized work instruction for Echo /SDC

f) There are many cardiac patients admitted in other general wards waiting for echo.

**PEOPLE**

a) There are few number of nursing staff on duty for pottering patients

b) There are few cardiologist to screen patients

c) There are few number of personnel trained for echo 

d) No one is allocated to take patients for echo procedure

e) Delay in patient's health seeking behavior

f) Delay in decision making

**MACHINE**

a) Few portable Echo machines.

b) Few portable oxygen cylinders to move patients on oxygen

c) There is only one Pediatrics Echo machine

