---
title: "Secialized Surgery Department"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warnings= FALSE,
                      message = FALSE)
library(tidyverse)
```

```{r}
data_11<- data.frame(quiz1a = c(rep(1,22),rep(2,4)),
                   quiz1b = c(rep(1,14),rep(2,12)),
                   quiz1c = c(rep(1,24),rep(2,2)),
                   quiz1d = c(rep(1,17),rep(2,9)),
                   quiz1e = c(rep(1,20),rep(2,6)),
                   quiz1f = c(rep(1,24),rep(2,2)),
                   quiz2a = c(rep(1,21),rep(2,5)),
                   quiz2b = c(rep(1,21),rep(2,5)),
                   quiz2c = c(rep(1,25),rep(2,1)),
                   quiz2d = c(rep(1,25),rep(2,1)),
                   quiz3a = c(rep(1,24),rep(2,2)),
                   quiz3b = c(rep(1,26),rep(2,0)),
                   quiz4a = c(rep(1,26),rep(2,0)),
                   quiz4b = c(rep(1,26),rep(2,0)),
                   quiz5a = c(rep(1,16),rep(2,10)),
                   quiz5b = c(rep(1,21),rep(2,5)))
```

```{r}
data_11 %>%
  gather(key="Questions",value, quiz1a:quiz1f) -> p
pp<-subset(p, select = c('Questions' , 'value'))
pp$Questions<-"people"
pp<-pp[pp$value==1,]

data_11 %>%
  gather(key="Questions",value, quiz2a:quiz2d) -> h
hh<-subset(h, select = c('Questions' , 'value'))
hh$Questions<-"people"
hh<-hh[hh$value==1,]

data_11 %>%
  gather(key="Questions",value, quiz3a:quiz3b) -> i
ii<-subset(i, select = c('Questions' , 'value'))
ii$Questions<-"environment"
ii<-ii[ii$value==1,]

data_11 %>%
  gather(key="Questions",value, quiz4a:quiz4b) -> j
jj<-subset(j, select = c('Questions' , 'value'))
jj$Questions<-"equipments"
jj<-jj[jj$value==1,]

data_11 %>%
  gather(key="Questions",value, quiz5a:quiz5b) -> k
kk<-subset(k, select = c('Questions' , 'value'))
kk$Questions<-"methods"
kk<-kk[kk$value==1,]

n<-rbind(pp,hh,ii,jj,kk)



ggplot(data=n,aes(x=factor(Questions),y=prop.table(stat(count)),
                  fill = factor(Questions),
                  label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  scale_y_continuous(labels = scales::percent)+
  geom_text(stat = 'count',
            vjust= -0.5,
            size= 3)+
   scale_fill_brewer(palette = "Dark2")+
  #scale_x_discrete( labels =c("Yes","No"))+
  labs(x = "Causes of delay",y = "Percentage Proportion",title = "Reduction of turnaround time of specialized \n surgery elective admissions.")+
  theme(legend.position = "none",plot.title = element_text(face="bold"))
```
<br> 

Analysis of data collected on some of the issues the pediatrics Department had identified as causes of delay in echo process had the following results.


**STAFF**

a) Ward round precedes admission process leading to delay. M YES

b) No one is allocated to take care of the patients coming for admission to the wards? (nurses/doctors)

c) Admission decision made by senior registrars/consultants? MYES

d) Discharge of patients determines number of patients to be admitted. []

e) Files take long to be brought to the ward? 

f) There's shortage of staff?


**PATIENTS**

Patients are not well versed with service points?


Directions to the various service points not clear

There are long ques at the service delivery points.

 Many scattered delivery points? 

**ENVIRONMENT**

Slow ICT systems

 Long queues at service delivery points

**EQUIPMENTS**

A. Inadequate beds? 

B. Squeezed space.


METHODS

1. Many people involved in the admission process/procedure?

2. Complex procedure/process Protocols (bureaucracy)
