---
title: "Reduction of Surgical site infection"
subtitle: Exploratory Data Analysis
output:
  word_document:
    toc: no
  html_document:
    df_print: default
    highlight: tango
    theme: readable
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
       echo = FALSE,
       message = FALSE,
       warning = FALSE
      )
library(tidyverse)
```

```{r}
data<-read.csv("C:/Users/adm/Desktop/HCQ/Reduction of surgical site infctn.csv")
```

## Gender
```{r}
X=sum(data$Gender==1)
Y=sum(data$Gender==2)
```

```{r}
ggplot(data,aes(x=factor(Gender),fill=factor(Gender)))+
  geom_bar()+
  theme_classic()+
  labs(x="Gender",title="GENDER DISTRIBUTION OF PATIENTS")+
  theme(legend.position = "none",plot.title = element_text(color = "blue",face = "bold"))+
  scale_fill_viridis_d()+
  scale_x_discrete(labels=c("Male","Female"))
  
```

From the above plots more males account for the abdominal surgeries in comparison to females.

## Type of surgery

```{r}
ggplot(data = data,aes(x=factor(Typeofsurgery),y=prop.table(stat(count)),
                         fill=factor(Typeofsurgery),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   labs(x="Type of Surgery",y="Percentage Frequency",title="TYPE OF SURGERY",fill="Type of Surgery")+
   theme(legend.position = "none",plot.title = element_text(face = "bold"))+
  scale_fill_discrete()+
   scale_x_discrete(labels=c("Elective","Emergency"))
```

<br>

Looking at the bar plots above  a majority of patients were taken in for an `emergency` surgery.This accounted for **93%**. Only **7%**of the patients had an `elective` surgery.


## Where the pre-operative care was done.

```{r}
ggplot(data = data,aes(x=factor(Wherepreoperative),
                       y=prop.table(stat(count)),
                         fill=factor(Wherepreoperative),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_viridis_d(begin=0, end = .95,option='D')+
   scale_x_discrete(labels=c("Ward","A & E"))+
   labs(x="Where preoperative care was done",y="Percentage Frequency",title="WHERE PREOPERATIVE CARE WAS DONE")+
   theme(legend.position = "none",plot.title = element_text(face = "bold"))
   
```
From the above graph `71% ` of pre-operative care was done in the *wards* and `29%` in  *Accidents and emergency* .

## Surgical Technique

```{r}
ggplot(data = data,aes(x=factor(Surgicaltech),
                       y=prop.table(stat(count)),
                         fill=factor(Surgicaltech),
                         label=scales::percent(prop.table(stat(count)))))+
  geom_bar(position = "dodge")+
  geom_text(stat = 'count',
            position = position_dodge(.9),
            vjust= -0.5,
            size= 3)+
  scale_y_continuous(labels = scales::percent)+
   scale_fill_viridis_d(begin=0, end = .95,option='D')+
   scale_x_discrete(labels=c("Laparatomy","Laparascopy"))+
   labs(x="Where preoperative care was done",y="Percentage Frequency",title="Surgical Technique")+
   theme(legend.position = "none",plot.title = element_text(face = "bold"))

```

For all the patients the surgical techniques undertaken were laparotomy.

<br>

## Was Prophylatic antibiotic given?

```{r}
data_1<-data.frame(druggvn=c("yes","no"),values =c(14,0))

ggplot(data = data_1,aes(x = factor(druggvn),y=prop.table(values),fill=factor(druggvn)))+
  geom_bar(stat = "identity")+
  geom_text(label = scales::percent(prop.table(data_1$values)),vjust = -0.1)+
  scale_y_continuous(labels=scales::percent)+
  theme(legend.position = 'none', plot.title = element_text(face="bold"))+
  labs(x="Was Prophylactic antibiotic given?", y= 'Percentage proportion',title="PROPHYLACTIC DRUG GIVEN OR NOT")
```

**100%** of the patients were given `prophylactic` drugs as seen from the above graph.

## What type of antimicrobial agent was given

```{r}
data_2 <- data.frame(agent =c("Cefazoline","Cefuroxime","Ceftriaxone","Vancomycin","Ceftazidime",
                              "Amoxicillin clavulanic acid","Metronidazole","Other (specify)"),values=c(13,0,1,0,0,0,0,0))

ggplot(data = data_2,aes(x = factor(agent),y=prop.table(values),fill=factor(agent)))+
  geom_bar(stat = "identity")+
  geom_text(label = scales::percent(prop.table(data_2$values)),vjust = -0.1)+
  scale_y_continuous(labels=scales::percent)+
  theme(legend.position = 'none',axis.text.x =  element_text(angle=45, vjust=1, hjust=1),plot.title = element_text(face="bold"))+
  labs(x="Type of antimicrobial agent ", y= 'Percentage proportion',title = "Distribution of Antimicrobial agent given")

```

92.9% of patients were administered with *cefazoline* with 7.1% being administered  *cetriaxone*.

