---
output:
  pdf_document:
    toc: yes
  htlm_document:
    toc: yes
    theme: readable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
       message = FALSE,
       warning = FALSE,
       fig.align='center')
```


# Chi-Square Test of Independence

The chi-square test of independence is used to analyze the frequency table (i.e. contingency table) formed by two categorical variables. The chi-square test evaluates whether there is a significant association between the categories of the two variables.

```{r}
 #Import the data
file_path <- "http://www.sthda.com/sthda/RDoc/data/housetasks.txt"
housetasks <- read.delim(file_path, row.names = 1)
head(housetasks)

```

```
The **chi-square test of independence** is used to analyze the frequency table (i.e. contengency table) formed by two categorical variables. The **chi-square** test evaluates whether there is a significant association between the categories of the two variables.
```

The data is a contingency table containing 13 housetasks and their distribution in the couple:
* **rows are the different tasks**
* **values are the frequencies of the tasks done :**
* **by the wife only**
* *alternatively*
* **by the husband only**
* **or jointly**

# Graphical display of contengency tables

Contingency table can be visualized using the function **balloonplot()** [in `gplots` package]. This function draws a graphical matrix where each cell contains a dot whose size reflects the relative magnitude of the corresponding component.

```{r}
library("gplots")
# 1. convert the data as a table
dt <- as.table(as.matrix(housetasks))
# 2. Graph
balloonplot(t(dt), main ="housetasks", xlab ="", ylab="",
            label = FALSE, show.margins = FALSE)
```

```
Note that, row and column sums are printed by default in the bottom and right margins, respectively. These values can be hidden using the argument show.margins = FALSE.
```

It’s also possible to visualize a contingency table as a mosaic plot. This is done using the function mosaicplot() from the built-in R package garphics:

```{r}
library("graphics")
mosaicplot(dt, shade = TRUE, las=2,
           main = "housetasks")
```

* The argument shade is used to color the graph

* The argument las = 2 produces vertical labels

<div class = "brown">
- **Note that the surface of an element of the mosaic reflects the relative magnitude of its value.**
</div>


* Blue color indicates that the observed value is higher than the expected value if the data were random

* Red color specifies that the observed value is lower than the expected value if the data were random

<style>
div.blue {background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">
- From this mosaic plot, it can be seen that the house tasks Laundry, Main_meal, Dinner and breakfast (blue color) are mainly done by the wife in our example. 
</div>

<br>

There is another package named vcd, which can be used to make a mosaic plot (function mosaic()) or an association plot (function `assoc()`).


# Chi-square test basics

Chi-square test examines whether rows and columns of a contingency table are statistically significantly associated.

* **Null hypothesis (H0):** the row and the column variables of the contingency table are independent.

* **Alternative hypothesis (H1):** row and column variables are dependent
For each cell of the table, we have to calculate the expected value under null hypothesis.

```{r}
chisq <- chisq.test(housetasks)
chisq
```


<div class = "brown">
- In our example, the row and the column variables are statistically significantly associated (p-value = 0).
</div>

The observed and the expected counts can be extracted from the result of the test as follows:

```{r}
# Observed counts
chisq$observed
```

```{r}
# Expected counts
round(chisq$expected,2)
```

